Real Time Genome Scan Using GPU

### 2019-07-10 paper.tex
adding GPU part in result

@@

\subsection{Benefit of using GPU}
To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
It includes the following steps (step 4 is optional):
\begin{itemize}
	\item Calculate standadized matrix for input matrices.
	\item Get correlation matrix
	\item Caclulate LOD score
	\item Find maximum LOD score for each phenotype. 
\end{itemize}

We profiled the entire genome scan program to get a complete picture of the execution process.
Our profiling result shows that 90\% of runtime is spent on calculating correlation matrix and LOD score. 
Since both involve matrix operations (multiplication and elementwise operations respectively) which are highly parallelizable, we used GPU to speed up matrix operations.

Table \ref{tab:speedup-table} shows timing between CPU and GPU

		\end{tabularx}
		\label{tab:speedup-table}
	\end{tableminipage}
\end{table*}

@@

 {\em GPUs vs CPUs:} GPUs are attractive since they provide a
massive number of cores at a lower price range.
Matrix multiplication and element-wise operations are amendable
to GPU's heterogeneous computing architecture, since both have
no data race conditions and low data dependencies.
\begin{figure}[!htb]
	\centering
	\caption{Variation of GPU vs CPU speedup with matrix shape
	}
	\label{GPUCPUShape}
	\includegraphics[scale = 0.36]{figs/speedup.png}
\end{figure}

@@

>>>>

		\end{tabularx}
		\footnotetext[1]{There are overhead of using GPU, such as data transfer and setting up context, the timing shown for CPU \& GPU here included all overhead for fairness. We ran the genome scan process 10 times and choose the median to remove randomness of each run, and the warm up time of GPU.}
		\label{tab:speedup-table}
	\end{tableminipage}
\end{table*}

@@

\subsection{Benefit of using GPU}
In parallel computing, Amdahl's law is used to find the theoretical maximum speedup by improving a particular part of a program.  
For example, if a program took ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, then, the theoretical speedup, no matter how many processors are used, can not be more than ten times, because the minimum execution time of this program is one minute. 
Therefore, profiling the entire genome scan process is the prerequisit in order to begin the optimization. 
Often, profiling would consider space and time complexity. 
Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
We used Julia's built-in Sampling Profiler (%TODO reference here.)
to find our target function for GPU because it is less intrusive compared to other profiling methods.

The genome scan process includes the following steps:
\begin{itemize}
	\item Calculate standadized matrix for input matrices.
	\item Get correlation matrix
	\item Caclulate LOD score
\end{itemize}
Our profiling result shows that step two and three takes up 90\% of time, and both involve parallelizable matrix operations. 
So they are our candidates for GPU accleration.
To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of time. 
In contrast, it is slower, whether using single or double precision for the data type.
We used a GPU profiler \textit{nvprof} %TODO reference here
to investigate what is the bottleneck of GPU. 
Result is, 98\% of time is spent on data transfer from GPU to CPU (device to host).
As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
For BXD spleen dataset, the Y' and G matrix is about 40MB %TODO: double check on this number
, while R matrix is about 4GB.
For our test hardware Nvidia TESLA V100, data transfer rate from CPU to GPU (host to device) is 4.3GB/s, and GPU to CPU (device to host) is 2.2GB/s. %TODO: double check the numbers here. 
Larger data size at a slower transfer rate makes GPU seems unattractive for computing genome scan, at least for BXD dataset.

To cope with this limitaion of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
Compared to CPU only, which took 1.77 seconds, this method gains 22 times speedup.

\begin{figure}[!htb]
	\centering
	\caption{Variation of GPU vs CPU speedup with matrix shape
	}
	\label{GPUCPUShape}
	\includegraphics[scale = 0.36]{figs/speedup.png}
\end{figure}

\iffalse
We profiled the entire genome scan program to get a complete picture of the execution process.
Our profiling result shows that 90\% of runtime is spent on calculating correlation matrix and LOD score. 
Since both involve matrix operations (multiplication and elementwise operations respectively) which are highly parallelizable, we used GPU to speed up matrix operations.

@@

 {\em GPUs vs CPUs:} GPUs are attractive since they provide a
massive number of cores at a lower price range.
Matrix multiplication and element-wise operations are amendable
to GPU's heterogeneous computing architecture, since both have
no data race conditions and low data dependencies.

@@

\fi
